---
title: "Reactive Cysteine Prediction: Exploratory Data Analysis"
author: "Bryn Reimer"
date: "27 July 2022"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(broom) # for PCA, tidy style
library(ggplot2)
library(ggbeeswarm)
library(ROCR) # for receiver operating curve
library(PRROC) # for precision-recall curve, better for imbalanced data
library(Peptides) # for generating ST-scale and T-scale

knitr::opts_chunk$set(echo=TRUE)
theme_set(theme_minimal())
```

## Exploratory data analysis

First, we read in the data from *pdbs_cys_allinfo.csv*. We examine each of the variables 
to determine their type and usefulness for modeling.

```{r read_data, include=FALSE}
dat1 <- read.csv("../v1/pdbs_cys_allinfo.csv")
odat2 <- read.csv("../v2/pdbs_cys_allinfo.csv")
dat1d <- dat1[!duplicated(dat1[,c("Res","Uniprot.Name")]),]
dat2d <- dat2[!duplicated(dat2[,c("Res","Uniprot.Name")]),]

dat_rosetta <- read.csv("../rosetta/pdbs_cys_allinfo_cutoff_none.csv")
dat_rosetta7 <- read.csv("../rosetta/pdbs_cys_allinfo_cutoff_7A.csv")
dat_rosetta8 <- read.csv("../rosetta/pdbs_cys_allinfo_cutoff_8A.csv")
dat_rosetta9 <- read.csv("../rosetta/pdbs_cys_allinfo_cutoff_9A.csv")
dat_rosetta10 <- read.csv("../rosetta/pdbs_cys_allinfo_cutoff_10A.csv")

#v1_prots <- unique(paste0(dat1$Uniprot.ID,"_", dat1$Pos))
#v2_prots <- unique(paste0(dat2$Uniprot.ID,"_", dat2$Pos))

#dat1$ids <- paste0(dat1$Uniprot.ID,"_", dat1$Pos)
#dat2$ids <- paste0(dat2$Uniprot.ID,"_", dat2$Pos)

#dat1_dedup <- dat1[-which(duplicated(dat1$ids)),]
#dat2_dedup <- dat2[-which(duplicated(dat2$ids)),]

#rownames(dat1_dedup) <- dat1_dedup$ids
#rownames(dat2_dedup) <- dat2_dedup$ids

#dat1_dedup <- dat1_dedup[order(dat1_dedup$ids),c("X.Exposure", "pKa","ligand.type")]
#dat2_dedup <- dat2_dedup[order(dat2_dedup$ids),c("X.Exposure", "pKa","ligand.type")]

#in.common <- intersect(rownames(dat1_dedup), rownames(dat2_dedup))

#plot(dat1_dedup[in.common, "pKa"] ~ dat2_dedup[in.common, "pKa"])

#plot(dat1[v1_prots, "X.Exposure"] ~ dat2[v1_prots, "X.Exposure"])

## helper function
parse_residues <- function(res_string) {
  res_ids <- lapply(strsplit(res_string, "|", fixed=T), function(x){gsub(".*_", "", x)})
  res_names <- lapply(res_ids, function(x){substr(x, 1, 1)})
  return(res_names)
}

```

```{r describe_feats, include=TRUE}
head(table(dat$Res))
table(dat$Type) # all are Polar
hist(dat$X.Exposure)
hist(dat$S.A.)
hist(dat$pKa)
hist(dat$Charge)
# ignore NearRes for now -- how calculated?
head(table(dat$Uniprot.ID))
head(table(dat$Uniprot.Name))
head(table(dat$CYS.link))
head(table(dat$ligand.link))
head(table(dat$ligand.type))
```

Our goal is to predict ligand.type; either "C_or_S" (reactive) or "free" (nonreactive). To do this, it is useful to subset the full dataset, which includes cysteines that are conjugated with zinc atoms, to only those cysteines which are either bound to a ligand ("C_or_S") or which are "free".

In addition, we take this time to sanitize our features a little bit. We log-transform the exposure (solvent exposure) variable since we are most interested in changes between, for example, 0 and 10\% exposure than we are in changes between 70\% and 80\% exposure. We take the nearby residues ("NearRes") and transform them into a sequence of amino acids which are passed into the [T-scale](https://www.sciencedirect.com/science/article/abs/pii/S0022286006006314) and [ST-scale](https://pubmed.ncbi.nlm.nih.gov/19373543/) algorithms using the [Peptides package](https://rdrr.io/cran/Peptides/). We are left with 8 ST-scale components and 5 T-scale components.

```{r process_data, include=FALSE}

all_dats <- list(dat1, dat2, dat1d, dat2d,
                 dat_rosetta, dat_rosetta7,
                 dat_rosetta8, dat_rosetta9,
                 dat_rosetta10)
dat.names <- c("dat1","dat2","dat1d","dat2d",
               "dat_rosetta","dat_rosetta7",
               "dat_rosetta8","dat_rosetta9",
               "dat_rosetta10")
names(all_dats) <- dat.names

for (name in dat.names) {
  dat <- all_dats[[name]]
  print(name)
  subset_dat <- dat[which(dat$ligand.type != "other"),]
  print(dim(subset_dat))
  
  df <- data.frame(y=as.factor(c(1,0)[as.factor(subset_dat$ligand.type)]))
  df$exposure <- subset_dat$X.Exposure
  df$log_exposure <- log(df$exposure + 1)
  df$res <- subset_dat$Res
  #df$charge <- as.factor(round(subset_dat$Charge))
  df$pdbid <- as.factor(substr(subset_dat$PDBID.chain, 0,4))
  df$chain <- as.factor(substr(subset_dat$PDBID.chain, 6, 6))
  df$pka <- subset_dat$pKa
  df$gene <- as.factor(subset_dat$Uniprot.Name)
  df$near_res <- parse_residues(subset_dat$NearRes)
  st_scales <- stScales(sapply(df$near_res, function(x){paste0(x[which(x!="X")], collapse="")}))
  t_scales <- tScales(sapply(df$near_res, function(x){paste0(x[which(x!="X")], collapse="")}))
  
  df$catalytic <- sapply(df$near_res, function(x) {
    grepl("HN", paste0(x, collapse=""))
  })
  df$has_h <- sapply(df$near_res, function(x) {
    grepl("H", paste0(x, collapse=""))
  })
  
  df$st1 <- unlist(st_scales)[seq(1, length(unlist(st_scales)), by=8)]
  df$st2 <- unlist(st_scales)[seq(2, length(unlist(st_scales)), by=8)]
  df$st3 <- unlist(st_scales)[seq(3, length(unlist(st_scales)), by=8)]
  df$st4 <- unlist(st_scales)[seq(4, length(unlist(st_scales)), by=8)]
  df$st5 <- unlist(st_scales)[seq(5, length(unlist(st_scales)), by=8)]
  df$st6 <- unlist(st_scales)[seq(6, length(unlist(st_scales)), by=8)]
  df$st7 <- unlist(st_scales)[seq(7, length(unlist(st_scales)), by=8)]
  df$st8 <- unlist(st_scales)[seq(8, length(unlist(st_scales)), by=8)]
  
  df$t1 <- unlist(t_scales)[seq(1, length(unlist(t_scales)), by=5)]
  df$t2 <- unlist(t_scales)[seq(2, length(unlist(t_scales)), by=5)]
  df$t3 <- unlist(t_scales)[seq(3, length(unlist(t_scales)), by=5)]
  df$t4 <- unlist(t_scales)[seq(4, length(unlist(t_scales)), by=5)]
  df$t5 <- unlist(t_scales)[seq(5, length(unlist(t_scales)), by=5)]
  
  saveRDS(df, paste0("cleaned_",name,".rds"))
}

# t scales and st scales are order-invariant
```

We can then explore the cleaned data frame and examine some preliminary plots to better understand the associations between our features of interest.

```{r explore_processed, include=TRUE}
glimpse(df)
continuous <-select_if(df, is.numeric)
summary(continuous)
# exposure will likely need to be transformed
# pka is roughly normally distributed

df %>% ggplot(aes(y, pka, color=y)) + 
  geom_quasirandom() + 
  geom_abline(slope=0, intercept=8.3) +
  theme_bw()

df %>% ggplot(aes(pka, color=y)) + 
  geom_density() + 
  ggtitle(paste0("Distribution of cov cys +/- examples")) +
  theme_bw()

df %>% ggplot(aes(log_exposure, color=y)) + 
  geom_density() + 
  ggtitle(paste0("Distribution of cov cys +/- examples")) +
  theme_bw()

df %>% 
  ggplot(aes(pka, log_exposure, color=y, alpha=has_h)) +
  geom_point(size=2) +
  ggtitle("Catalytic triad CHN") + 
  theme_bw()

```

Using dimensionality reduction (PCA), we can look at whether we have sufficient signal to create a discriminative linear model. The PCA below shows excellent signal, which means we are likely to be able to build a predictive model.

```{r run_pca, echo=FALSE}
pca_fit <- df %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  prcomp()

pca_fit %>% 
  augment(df) %>%
  ggplot(aes(.fittedPC1, .fittedPC2, color=y)) +
  geom_point(size=1.5) +
  theme_minimal()

#saveRDS(df, "cleaned_cyscov_data.rds")
```


Finally, we can examine individual genes of interest. Here, only those with at least 10 cysteines are shown.

```{r show_individual, echo=FALSE}
for (gene in unique(df$gene)) {
  tmp_subset <- df[which(df$gene == gene),]
  if (nrow(tmp_subset) >= 10) {
    g <- ggplot(tmp_subset, aes(pka, log_exposure, color=y)) +
      geom_point(size=1.5) +
      xlim(0, 19) + 
      ylim(0, 5) +
      ggtitle(gene) +
      theme_bw()
    print(g)
  }
}
```







